CommonMind Consortium: DLPFC

### 2020-09-22 combined_analysis/combine_all_DLPFC.Rmd
freeze

---
title: "CommonMind Consortium: DLPFC"
subtitle: "Analysis of read count data by cohort"
author:
- name: "[Gabriel Hoffman](http://gabrielhoffman.github.io)$^{1}$ and Kelsey Montgomery$^{2}$"
  affiliation: | 
    [1] Icahn School of Medicine at Mount Sinai, New York
    [2] Sage Bionetworks, Seattle, WA
date: "Run on `r Sys.time()`"
documentclass: article
output:
  html_document:
    toc: true
    toc_float: true
params:
  upload: TRUE
---

@@

knitr::opts_chunk$set(
  echo = FALSE,
  warning=FALSE,
  message=FALSE,
  error = FALSE,
  tidy = FALSE,
  cache = TRUE)
```

@@

form = ~ ageOfDeath:Institution + Reported_Gender + Dx + Institution + Dx

@@

vp = fitExtractVarPartModel( t(Y), form, METADATA, BPPARAM=SerialParam())
plotPercentBars(vp) + ggtitle("Explain cell type composition")

ymax = max(rowSums(vp[,-ncol(vp)]))*110
plotPercentBars(vp) + scale_y_continuous(expand=c(0,0), limits=c(0, ymax))
```

```{r sex.graph, fig.width=7, fig.height=7}
library(ggplot2)
library(ggraph)
library(tidygraph)

df_cca2 = df_cca[with(df_cca, v1!=v2),]

@@

@@

## Plot top examples
```{r plot.forrest.dlpfc.Sex, fig.width=8, fig.height=20}  
figList = lapply(df_meta[[coef]]$Symbol[1:10], function(geneSymbol){

@@

@@

@@

@@

# Dx test with in Males and then Females

>>>>

---
title: "CommonMind Consortium: DLPFC"
subtitle: "Analysis of read count data by cohort"
author:
- name: "[Gabriel Hoffman](http://gabrielhoffman.github.io)$^{1}$ and Kelsey Montgomery$^{2}$"
  affiliation: | 
    [1] Icahn School of Medicine at Mount Sinai, New York
    [2] Sage Bionetworks, Seattle, WA
date: "Run on `r Sys.time()`"
documentclass: article
output:
  html_document:
    toc: true
    toc_float: true
params:
  upload: FALSE
---

@@

knitr::opts_chunk$set(
  echo = FALSE,
  warning=FALSE,
  message=FALSE,
  error = FALSE,
  tidy = FALSE,
  dev = c('png', 'pdf'),
  cache = TRUE)

colorArray = colors()[grep("grey|gray", colors(), invert=TRUE)]

@@

form = ~ ageOfDeath:Institution + Reported_Gender + Dx + Institution

@@

df = lapply( colnames(Y), function(id){

  fit = lm(Y[,id] ~ ageOfDeath:Institution + Reported_Gender + Dx + Institution, METADATA)

  data.frame( Test     = rownames(anova(fit))[1:4],
              CellType = id, 
              p        = anova(fit)[,'Pr(>F)'][1:4], 
              frac     =  calcVarPart(fit)[1:4])
})
df = data.table(do.call(rbind, df))

df$Test[df$Test=='Reported_Gender'] = "Sex"
df$Test[df$Test=='ageOfDeath:Institution'] = "Age of death"
df$p.adj = p.adjust(df$p, "bonferroni")

ymax = max(df[,sum(frac),by="CellType"]$V1)*105
ggplot(df, aes(CellType, frac*100, fill=Test)) + geom_bar(stat="identity") + coord_flip() + theme_bw() + theme(aspect.ratio=1, panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank()) +  ylab("Variance explained (%)") + theme(axis.line = element_line(colour = "transparent"), 
        axis.line.x = element_line(colour = "black"), panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), panel.border = element_blank(), 
        panel.background = element_blank(), axis.ticks.y = element_blank(), 
        legend.key = element_blank(), plot.margin = unit(c(0, 
            0.3, 0, 0.8), "cm")) + scale_y_continuous(expand=c(0,0), limits=c(0, ymax)) + xlab('')

ggplot(df, aes(CellType, Test, fill=-log10(pmax(1e-16,p)), label=ifelse(p.adj<0.05, '*', ''))) + geom_tile() + theme_bw() + scale_fill_gradient(name = bquote(-log[10]~P), low="white", high="red", limit=c(0, NA)) + theme(aspect.ratio=1, panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank()) + geom_text() + coord_flip()
```

```{r sex.graph, fig.width=7, fig.height=7, dev="png"}
df_cca2 = df_cca[with(df_cca, v1!=v2),]

@@

```{r plot.gene_set_test.marginal, eval=TRUE}
 plot_zenith_results = function( res, coef, Resource, FDR = 0.05){

  df = df[order(df$PValue, decreasing=TRUE),]
  df$Set = factor(df$Set, df$Set)
  ymax = max(-log10(df$PValue)) * 1.05
  ggplot(df, aes(Set, -log10(PValue))) + geom_bar(stat="identity", fill="navy") + theme_bw() + theme(aspect.ratio=1,plot.title = element_text(hjust = 0.5)) + coord_flip() + ylab(bquote(-log[10]~P)) + xlab('') + ggtitle(coef) + scale_y_continuous(expand=c(0,0), limits=c(0, ymax))
}

 plot_zenith_results( res_meta_zenith, "DxSCZ", "GO", FDR=0.05)
```

```{r, fig.width=9}
 plot_zenith_results( res_meta_zenith, "DxSCZ", "MSigDB", FDR=0.05)
```

@@

## Plot top examples - not chrY
```{r plot.forrest.dlpfc.Sex.autosomes, fig.width=8, fig.height=20}  
Symbols = df_meta[[coef]][df_meta[[coef]]$Chrom !='Y',]$Symbol[1:10]
figList = lapply(Symbols, function(geneSymbol){

## Plot top examples - autosomes
```{r plot.forrest.dlpfc.Sex, fig.width=8, fig.height=20}  
Symbols = df_meta[[coef]][df_meta[[coef]]$Chrom %in% 1:22,]$Symbol[1:10]

figList = lapply(Symbols, function(geneSymbol){

@@

```{r plot.gene_set_test.marginal.Sex, eval=TRUE}
 plot_zenith_results( res_meta_zenith, "Reported_GenderMale", "GO", FDR=0.05)         
 plot_zenith_results( res_meta_zenith, "Reported_GenderMale", "MSigDB", FDR=0.05)        
```

@@

```{r plot.gene_set_test.marginal.Sex.autosome, eval=TRUE}
 plot_zenith_results( res_meta_zenith.sex_autosome, "Reported_GenderMale", "GO", FDR=0.1) 
 plot_zenith_results( res_meta_zenith.sex_autosome, "Reported_GenderMale", "MSigDB", FDR=0.1)   
```

@@

```{r plot.gene_set_test.marginal.Age, eval=TRUE}
 plot_zenith_results( res_meta_zenith, "ageOfDeath", "GO", FDR=0.01)   
 plot_zenith_results( res_meta_zenith, "ageOfDeath", "MSigDB", FDR=0.01)    
```

@@

```{r plot.gene_set_test.marginal.inter_sex.disease, eval=TRUE}
coef = 'DxSCZ:Reported_GenderMale'   
 plot_zenith_results( res_meta_zenith.inter_sex.disease, coef, "GO", FDR=0.1)  
 plot_zenith_results( res_meta_zenith.inter_sex.disease, coef, "MSigDB", FDR=0.1)

 plot_zenith_results( res_meta_zenith.inter_sex.disease_autosome, coef, "GO", FDR=0.1) 
 plot_zenith_results( res_meta_zenith.inter_sex.disease_autosome, coef, "MSigDB", FDR=0.1)    
```

# Dx test separately in Males and then Females
```{r dream.sex.disease.separately}
form.lst = list(
  `MSSM-Penn-Pitt`  = ~ Dx + RIN +scale(IntronicRate) + scale(IntragenicRate) + scale(IntergenicRate) + scale(rRNARate) + Institution*(ageOfDeath + cellFrac_ilr_1 + cellFrac_ilr_2 + cellFrac_ilr_3),
  `NIMH-HBCC`       = ~ Dx + RIN + scale(IntronicRate) + scale(IntragenicRate) + scale(IntergenicRate) + scale(rRNARate) + ageOfDeath + cellFrac_ilr_1 + cellFrac_ilr_2 + cellFrac_ilr_3)

# DE for males and females separately
fit.separ = lapply( vobj.lst, function(vobj){

  fit.lst = lapply(unique(METADATA$Reported_Gender), function(Sex){

    i = match(colnames(vobj), rownames(METADATA))
    info = METADATA[i,]
    idx = which(info$Reported_Gender == Sex)

    fit = dream(vobj[,idx], form_use, info[idx,], BPPARAM = SnowParam(18))  
    eBayes( fit )
  })
  names(fit.lst) = unique(METADATA$Reported_Gender)
  fit.lst
  })
names(fit.separ) = names(vobj.lst)

# meta-analysis
df_meta.separ = lapply(unique(METADATA$Reported_Gender), function(Sex){
  tab.lst = lapply( names(vobj.lst), function(chrt){
    topTable(fit.separ[[chrt]][[Sex]], coef="DxSCZ", sort.by='none', number=Inf)
  })

  res = run_meta_analysis( tab.lst, "FE") %>% getGeneSymbol
  res$t = with(res, logFC/se)
  res
})
names(df_meta.separ) = unique(METADATA$Reported_Gender)
```

```{r separ.concordance.scz, fig.height=6, fig.width=10}
# Concordance analysis
resConcord.separ = plot_concordance( df_meta.separ, showGenes = df_meta[['DxSCZ']]$Symbol[1:20])

main = "Concordance of SCZ across sexes"
plot_grid(  resConcord.separ$fig_logFC + ggtitle(main),
            resConcord.separ$fig_tstat + ggtitle(main), nrow=1)

data.frame(Cohort = names(df_meta.separ),
          "# DE"  = resConcord.separ$n.de,
          pi1     = resConcord.separ$pi_single,
          check.names=FALSE) %>% 
          mutate_if(is.numeric, format, digits=3) %>% kable(caption="pi1 within dataset") %>% kable_styling(full_width=FALSE)

resConcord.separ$tab_pi %>% mutate_if(is.numeric, format, digits=3) %>% kable(caption="pi1 for replication") %>% kable_styling(full_width=FALSE)
resConcord.separ$tab_corr %>% mutate_if(is.numeric, format, digits=3) %>% kable(caption="Correlation between t-statistics") %>% kable_styling(full_width=FALSE)
```

```{r compare.separate, fig.width=5, fig.height=5}
df = merge( df_meta.separ[[1]], df_meta.separ[[2]], by="row.names")

fit = with(df, lm(logFC.y ~ logFC.x))

coef(summary(fit) ) %>%  kable(caption="logFC is larger in males") %>% kable_styling(full_width=FALSE)

 # compare logFC
tstat = with(df, (logFC.y - logFC.x) / (sqrt(se.y^2 + se.x^2)))

qqt(tstat)
abline(0, 1, col="red")
p = 2*pt(abs(tstat), df=778, lower.tail=FALSE)
table(p.adjust(p) < 0.05)
```

## Plot top examples
```{r plot.forrest.dlpfc.separ, fig.width=8, fig.height=40}

figList = lapply(df_meta.separ[[1]]$Symbol[1:20], function(geneSymbol){
  fig1 = plot_forrest( df_meta.separ, geneSymbol, plot2='p')

  plot_grid(fig1, nrow=1)
})
